### 2015-05-20 doc/thesis/typedlua.tex
Improving the text about recursive functions and the differences between any and value

@@

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an unannotated return type.
This means that we can rewrite the previous example as follows:
\begin{verbatim}
    local function succ (n:integer)
      return n + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

Typed Lua assigns the type \texttt{any} to the unannotated declarations
that it does not infer a more specific type.
More precisely, Typed Lua does not infer more specific types to input
parameters of function declarations and to the return type of
recursive functions.

@@

This example compiles without any warnings because we annotated
the return type of \texttt{factorial}.

@@

@@

As we can see in Figure \ref{fig:functions}, second-level types consist
of either \emph{tupletype} or \emph{rettype}.
A second-level type is either the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
The type void \texttt{()} represents the type of the empty tuple.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuple types appear in the return type of function types to
represent overloading on the return type.
We will explain union types in more detail in the next section.
We can use only one first-level type \texttt{t} in the return type
because it is syntactic sugar to the tuple type \texttt{(t)}.

>>>>

Before we start discussing examples of Typed Lua code, it is worth
mentioning that there is a subtle difference between the dynamic
type \texttt{any} and the top type \texttt{value}.
Although both types mean that they accept a value of any other type,
the type \texttt{value} is not a good option for handling the
interaction between dynamically typed and statically typed code.
Gradual typing uses the dynamic type \texttt{any} to identify
where it should insert run-time checks for asserting that dynamically
typed code does not violate statically typed code.
Typed Lua also uses the dynamic type \texttt{any} in this sense,
though it is an optional type system.
More precisely, we use \texttt{any} instead of \texttt{value} to allow
programmers blending dynamic and static typing because we use the
consistent-subtyping relation to formalize our optional type system,
as it is a first step to switch Typed Lua from optional typing to
gradual typing in the future.

@@

Typed Lua uses local type inference to assign more specific types to
some unannotated declarations.
More precisely, Typed Lua can infer the type of local variables and
the return type of local functions that are not recursive.
The inference that we implement in Typed Lua is quite simple, as it
uses only the type of the local expression.
For local variables, Typed Lua uses the type of the initialization
expression to assign a more specific type to an unannotated local variable.
For local functions, Typed Lua uses the type of the returned expression
to assign a more specific type to an unannotated return type.

This means that we can rewrite the previous example to use local type
inference for inferring the return type of \texttt{succ} and also for
inferring the type of \texttt{x}:
\begin{verbatim}
    local function succ (n:integer)
      return n + 1
    end
    local x = 7
    x = succ(x)
    print(x)      --> 8
\end{verbatim}

@@

Typed Lua assigns the dynamic type \texttt{any} to the unannotated declarations
that it does not infer a more specific type.
More precisely, Typed Lua does not infer more specific types to the input
parameters of function declarations and to the return type of
recursive functions.
The Typed Lua compiler cannot infer them because it performs type checking in
a single step that simulates the program execution.
We could have split type checking into two steps to try solving this
limitation, but it would have implications in the mechanisms
that Typed Lua uses to handle the discrimination of union types
and the refinement of table types.
We will discuss these features in Section \ref{sec:unions} and
Section \ref{sec:tables}, respectivelly.

@@

This example compiles without any warnings because we annotated
the return type of \texttt{factorial}.
Local type inference cannot use the type of the returned
expression when it includes a function call to the function that is
being type checked, as its return type is still unknown to the type checker.
For this reason, we need to annotate the return type of
recursive local functions to inform the type cheker what type
it should use while type checking their body.

@@

Even though Typed Lua can type check recursive functions when we
annotate their return type, it has some limitations for type checking
mutually recursive functions, even if we annotate their return type.
We will use the following example for discussing this limitation and
also for presenting an alternative solution:
\begin{verbatim}
    local even, odd

    function even (n:integer):boolean
      if n == 0 then return true
      elseif n > 0 then return odd(n - 1)
      else return odd(n + 1)
      end
    end

    function odd (n:integer):boolean
      if n == 0 then return false
      elseif n > 0 then return even(n - 1)
      else return even(n + 1)
      end
    end
\end{verbatim}

This example shows an attempt of annotating a common idiom that Lua
programmers use for defining mutually recursive functions, but it
generates compile-time warnings.
The problem is related to the fact that Lua does not split a program
into declarations and statements, and also to the fact that Typed Lua
performs type checking in a single step.
This means that forwarding the declaration of a local variable assigns
the type \texttt{any} to it, making the compiler warn the programmers
about function calls to this variable, as they can assign any value
to a variable that has the dynamic type.
In this example, the Typed Lua compiler generates warnings when we
call \texttt{even} and \texttt{odd}.

One way to overcome these warnings is to predeclare these functions
with an empty body before the actual declaration.
Even though this is a verbose solution and the Typed Lua compiler
has an option to suppress warnings related to the dynamic type
\texttt{any}, it ensures that further assignments to forwarded
local variables will not change their type.
Next we show how we can apply this alteranative solution to the
previous example, making it compile without any warnings:
\begin{verbatim}
    local function even (n:integer):boolean return true end
    local function odd (n:integer):boolean return false end

    function even (n:integer):boolean
      if n == 0 then return true
      elseif n > 0 then return odd(n - 1)
      else return odd(n + 1)
      end
    end

    function odd (n:integer):boolean
      if n == 0 then return false
      elseif n > 0 then return even(n - 1)
      else return even(n + 1)
      end
    end
\end{verbatim}

@@

Figure \ref{fig:functions} shows that Typed Lua uses second-level
types in the definition of function types.
A second-level type is either the type void \texttt{()},
a tuple of first-level types optionally ending in a variadic type,
or a union of these tuples.
The type void \texttt{()} represents the type of the empty tuple.
A variadic type \texttt{t*} is a generator for a sequence of values
of the union type \texttt{t|nil}.
Union of tuple types appear in the return type of function types to
represent overloading on the return type.
We will explain union types in more detail in the next section.
We can use only one first-level type \texttt{t} in the return type
because it is syntactic sugar to the tuple type \texttt{(t)}.
